Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-20 Manuscript/manuscript.tex
Joe - retouche petit up to results (not-included)

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{This could be maybe extended.}
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. The detailed results from simple model systems can be then
used to understand the role of ions in complex biological systems.

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered higly relevant issue since the early days of
lipid bilayer simulations.
%Some pioneering simulation studies scaled
%the partial charges of atoms by one half to effectively include 
%polarizability \cite{jonsson86,egberts94}. Also, approaches that
%explicitly include polarizability has been introduced, but
%this has turned out to be practically complicated \cite{chowdhary13}.
In this work we circumvent the rather demanding explicit inclusion
of electronic polariization effects \cite{chowdhary13} by implicitly including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}. 
However, the concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13} both for derivation and application.

@@

While the scaling factor of $f_q = 0.75$ for ions in water improves their performance and is physically justifiable within the MDECC theory \cite{Leontyev2011},
it is not clear how the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of molecules, atomic partial charges within each molecule are not physical observables.
There are several schemes for the assignment of partial charges for biomolecules. \cite{Hu2007}
Currently, the most commonly employed scheme is RESP~\cite{RESP_paper,Singh1984}, 
which aims to reproduce the electrostatic potential obtained from quantum mechanical calculations around an ensemble the target molecules. 
The fixed partial atomic charges obtained by this method is hence 
the best fit of these partial charges that reproduces the electrostatic potential around the provided target conformations. 
Such averaging is well suited for the application of ECC, 
which targets at improving electrostatic interactions using a mean field correction. 
In practice, partial charges currently implemented in force fields may already 
include  to some extend some of the solvent electronic polarizability effects, i.e.,
the RESP charges are scaled to fit some experimental observable. 
Thus, we expect that the application of the ECC scaling factor, $f_q$, on the molecular partial charges
included in the available force fields do not necessarily require the integral, $f_q = \epsilon _{el} ^{-1/2}$, 
but instead it lies between 0.75 (no electronic polarizability in the charge calculation)
and $1$ (full electronic polarizability already included in the charge calculation).
%
% This is not really important for the paper:
%
%However, the RESP method has several remedies;  %and these are specific to RESP
%RESP charges show a dependence on the orientation and position
%and it does not provide numerically stable partial charges of burried atoms;~\cite{Hu2007}
%the RESP-derived values depend, for example, on the conformational
%sampling of the molecules and description of solvent 
%in the calculations~\cite{RESP_paper, Singh1984, Hu2007, maciejewski14}. 
%
%Hirshfeld-E partitioning: \cite{Verstraelen2013}
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~~\cite{RESP_paper,Singh1984}),
%
%

In this work, we produce a phospholipid model for classical MD simulations that
accurately describes the lipid head group response to varying concentrations
of monovalent and divalent cations. This is critical membrane feature, which is
poorly reproduced by currently available models, can affect not only membrane properties
in the presence of ions but also modulates the interaction with charged moieties in the surface.
Importantly, this response from simulations can be accurately
compared against experimental NMR data \cite{akutsu81,altenbach84,scherer89},
as discussed in Ref. \citenum{catte16} in section \ref{section:electrometer}.
To this end, we empirically explore the scaling factor parameter space, $f_q \in [0.75, 1.0]$ on the Lipid14~\cite{dickson14} force field.
We selected this force field as a starting point because their response to bound ions was apparently the most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding \todo{Do we really need to scale charges in non charged methylens in the choline and glycerol backbones? why? Does it really make a difference?}.
However, we do not modify the hydrocarbon chain parameters, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in most lipids models, including Lipid14 \cite{ollila16}.
This contrast with the behavior in glycerol backbone and head group
parameters which call for improvements in all available lipid models~\cite{botan15}.

Exploring different scaling factor values, we found out that ion binding and
related head group order parameter responses become weaker. 
The optimal behaviour of ion binding was observed with
the scaling factor $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
\todo{We should discuss how this can potentially affect the intermolecular interaction when mixing scaled and non scaled molecules}
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters with a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
again in agreement with experimental value (see Table~\ref{tab:apls}).

@@

\subsection{Salt concentrations and binding affinity}
Experimental works providing the lipid head group order
parameters report the salt concentrations in two different ways.
For DPPC, salt concentrations before solvating
the lipids are used~\cite{akutsu81}. Instead, for POPC,
the salt concentration of supernatant after solvation is
measured using atomic absorption~\cite{altenbach84}.
In this work, we will use the latter definition. For this we
will measure the salt concentration in the water bulk region,
\textit{i.e.}, the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study, Ref.~\citenum{catte16},
the concentration was calculated as in the DPPC experiments.
Despite of the measurable differences between two concentrations for CaCl$_2$ systems,
the qualitatively conclusions are not affected neither in this
nor in our previous work~\cite{catte16}.

To quantify the ion binding affinity to a membrane, we
calculated the relative surface excess of ions with respect to water $\Gamma_i^w$  \cite{chattorajBOOK}.
This quantity can be used to compare binding affinity between different models,
because it does not depend on the definition of the interfacial region
or the position of the Gibbs dividing plane between two bulk regions.
Therefore, we consider the hydrophobic interior of a membrane having zero concentrations of ions and water
as a bulk region, which is separated by a monolayer interface from 
a another region with bulk concentration of ions in water.
Since the boundary of the latter region is set to the edge of simulation box,
all molecules in simulation are within the defined interfacial region. 
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
This setup provides a simplified relation for $\Gamma_i^w$ in lipid bilayers simulations 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the water phase;
and $A_b$ is the size of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface in both leaflefts.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.5cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: Chemical structure of POPC with definitions of different order parameters calculated.
    Middle: Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13} {\color {red} X-label misses glycerol region}.
    Bottom: X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    }
\end{figure}

>>>>

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{This could be maybe extended.}
\todo{JOE: following doesn't sound right to me. The reasoning is weird.}
Since direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. The detailed results from simple model systems can be then
used to understand the role of ions in complex biological systems.

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered higly relevant issue since the early days of
lipid bilayer simulations.
%Some pioneering simulation studies scaled
%the partial charges of atoms by one half to effectively include 
%polarizability \cite{jonsson86,egberts94}. Also, approaches that
%explicitly include polarizability has been introduced, but
%this has turned out to be practically complicated \cite{chowdhary13}.
In this work we circumvent the rather demanding explicit inclusion
of electronic polariization effects \cite{chowdhary13} by implicitly including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
\todo{JOE: It would be a good idea to be consistent in using ECC resp. MDEC. I think that here shuld be MDEC as it cites Stuchebrukhov.}
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}. 
However, the concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11},
but significantly more simple than the explicit inclusion of electronic
polarizability \cite{chowdhary13} both for derivation and application.

@@

While the scaling factor of $f_q = 0.75$ for ions in water improves their performance and is physically justifiable within the MDEC theory \cite{Leontyev2011},
it is not clear how the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of molecules, atomic partial charges within each molecule are not physical observables.
There are several schemes for the assignment of partial charges for biomolecules. \cite{Hu2007}
Currently, the most commonly employed scheme is RESP~\cite{RESP_paper,Singh1984}, 
which aims to reproduce the electrostatic potential obtained from quantum mechanical calculations around an ensemble the target molecules. 
The fixed partial atomic charges obtained by this method are hence 
the best fit of pint charges reproducing the electrostatic potential around provided target conformations. 
Such a scheme is well suited for the application of ECC, 
which targets at improving electrostatic interactions using a mean field correction. 
In practice, partial charges currently implemented in force fields may already 
include  to some extend some of the solvent electronic polarizability effects, i.e.,
the RESP charges are scaled to fit some experimental observables. 
Thus, we expect that the application of the ECC scaling factor, $f_q$, on the molecular partial charges
included in the available force fields does not necessarily have to follow the relation $f_q = \epsilon _{el} ^{-1/2}$, 
but instead it lies between 0.75 (no electronic polarizability in the charge calculation)
and $1$ (full electronic polarizability already included in the charge calculation).
%
% This is not really important for the paper:
%
%However, the RESP method has several remedies;  %and these are specific to RESP
%RESP charges show a dependence on the orientation and position
%and it does not provide numerically stable partial charges of burried atoms;~\cite{Hu2007}
%the RESP-derived values depend, for example, on the conformational
%sampling of the molecules and description of solvent 
%in the calculations~\cite{RESP_paper, Singh1984, Hu2007, maciejewski14}. 
%
%Hirshfeld-E partitioning: \cite{Verstraelen2013}
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~~\cite{RESP_paper,Singh1984}),
%
%

In this work, we produce a phospholipid model for classical MD simulations that
accurately describes the lipid head group response to varying concentrations
of monovalent and divalent cations. This is a critical membrane feature, which is
poorly reproduced by currently available models; It affects not only membrane properties
in the presence of ions but also modulates the interaction with charged moieties in the surface.
Importantly, this response from simulations can be accurately
compared against experimental NMR data \cite{akutsu81,altenbach84,scherer89},
as discussed in Ref. \citenum{catte16} in section \ref{section:electrometer}.
To this end, we empirically explore the scaling factor parameter space, $f_q \in [0.75, 1.0]$ on the Lipid14~\cite{dickson14} force field.
We selected this force field as a starting point because their response to bound ions was apparently the most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding \todo{Do we really need to scale charges in non charged methylens in the choline and glycerol backbones? why? Does it really make a difference?
JOE: Cannot answer this without actually doing it. The selection was a matter of choice -- in the beginnig we argued with Samuli about this. I said that we shuld select the polar atoms in this selection and leave out the rest as ECC would mean little difference to them. Samuli then argued in a way that if it makes no difference, why not apply ECC onto such atoms too? Hence we applied ECC on all of the atoms.}.
However, we do not modify the hydrocarbon chain parameters, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in most lipids models, including Lipid14 \cite{ollila16}.
This contrasts with the behavior in glycerol backbone and head group order
parameters which call for improvements in all available lipid models~\cite{botan15}.

Exploring different scaling factor values, we found out that ion binding and
related head group order parameter responses become weaker. 
The optimal behaviour of ion binding was observed with
the scaling factor $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
\todo{JOE: following discussion shall be modified in the enlightenment of our recent ECC-discussions.}
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of the lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
\todo{We should discuss how this can potentially affect the intermolecular interaction when mixing scaled and non scaled molecules.
JOE: I think that we rather increasingly see that there's nothing like ``fully non-scaled'' with the exception of ions with integer charges. So the discussion shall be rather more about the interaction of our ``scaled'' (I'd still rather call it ECC-corrected or whatever) and ``semi-scaled'' models.}
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters with a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
again in agreement with experimental value (see Table~\ref{tab:apls}).

@@

\subsection{Salt concentrations and binding affinity}
Experimental works providing the lipid head group order
parameters report the salt concentrations in two different ways.
For DPPC, salt concentrations before solvating
the lipids are used~\cite{akutsu81}. Instead, for POPC,
the salt concentration of supernatant after solvation is
measured using atomic absorption~\cite{altenbach84}.
\todo{JOE: I'm afraid that the above description might sound to the reader like for some reasons one method is used with DPPC, but a different one with POPC. I think simply writing one way is\dots another way is \dots (or similarly) will be less confusing.}
In this work, we will use the latter definition. For this we
will measure the salt concentration in the water bulk region,
\textit{i.e.}, the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study, Ref.~\citenum{catte16},
the concentration was calculated as in the DPPC experiments.
Despite of the measurable differences between two concentrations for CaCl$_2$ systems,
the qualitatively conclusions are not affected neither in this
nor in our previous work~\cite{catte16}.

To quantify the ion binding affinity to a membrane, we
calculated the relative surface excess of ions with respect to water $\Gamma_i^w$  \cite{chattorajBOOK}.
This quantity can be used to compare binding affinity between different models,
because it does not depend on the definition of the interfacial region
or the position of the Gibbs dividing plane between two bulk regions.
\todo{JOE: following is very unclear to me. I think we can write it in a more concise way.}
Therefore, we consider the hydrophobic interior of a membrane having zero concentrations of ions and water
as a bulk region, which is separated by a monolayer interface from 
a another region with bulk concentration of ions in water.
Since the boundary of the latter region is set to the edge of simulation box,
all molecules in simulation are within the defined interfacial region. 
%our whole simulation box as
%an interface flanked by a water phase and an ideal phase with the characteristics of to
%the hydrophobic region in the middle of the membrane where no ions or waters are present.
This setup provides a simplified relation for $\Gamma_i^w$ in lipid bilayers simulations 
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where $n_w$ and $n_i$ are the total number of waters and ions in the system;
$C_w$ and $C_i$ are their respective bulk concentrations in the water phase;
and $A_b$ is the size of the box in the membrane plane.
The total area of the interface is twice the area of the membrane, $2A_b$,
because bilayers have an interface in both leaflets.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.5cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: Chemical structure of POPC with definitions of different order parameters calculated.
    Middle: Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.
    Bottom: X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    } 
    \todo{X-label misses glycerol region. JOE: glycerol belongs to the headgroup part, but, yes, it wouldn't hurt to write it explicitly.}
\end{figure}
